A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-10-12 main.tex
sec 4 updates - in progress

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered, which we refer to as ``fine'' and ``coarse''.  We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, representing a continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations (also known as founding populations) are of equal size. We note that the distinction between fine and coarse structure is an entirely different concept than the level of admixture in the simulated data. It is the size of the subpopulations, not the level of admixture, that determines what we refer to as fine or coarse structure. For example, both the Independent Subpopulation and 1D Linear Admixture simulated data may be comprised of fine or coarse structure. However, the subpopulations will be more distinct in the unadmixed Independent Subpopulation data than in the 1D Linear Admixed data. The Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \ref{fig:admixed} - \ref{fig:empirical} illustrate the simulated kinship structures, as well as that for the Empirical SNP data.

@@

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. This is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by nonoverlapping genetic subpopulations, meaning that a greater proportion of the environmental effect is in the column space of $\boldsymbol{P}_{\bX}$. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:sim_env_conf}. Here, we have CP $= 0.47, 0.99, 0.61$ for the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. \pb{let's move the 0.61 to the next paragraph} The large CP of the Independent Subpopulations model reflects the extreme confounding and estimation bias for this data, for which the PC-Lasso and LMM-Lasso methods are able to correct. Meanwhile, the less dramatic improvement in MSE for the 1D Linear Admixture data by the PC-Lasso and LMM-Lasso methods confirms what we know from this data's substantially smaller CP value; that there is less confounding and bias which can be corrected for.

Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical CP with those from the simulated data sets is not straightforward. We note that the CP decreases in the presence of subpopulations of unequal sizes and with more outbred subpopulations as in the Empirical data.

@@

Given the superior performance of PC-Lasso and LMM-Lasso compared to the ordinary Lasso in the presence of environmental confounding, we next compare these methods in the presence of fine and coarse subpopulation structures \pb{as defined in section....}. Phenotypes were simulated using a 1:1 SNR, $\eta = 0.5$, and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/mse_diff_subpops}
     \includegraphics[width=0.8\textwidth]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. \pb{This figure could maybe be shorter -- takes up a lot of vertical room that isn't really necessary}}
    \label{fig:big_vs_small}
\end{figure}

@@

In addition to evaluating PC-Lasso and LMM-Lasso in the presence of coarse and fine subpopulation structure, we also consider various patterns of environmental effects. Figure \ref{fig:big_small_gamma} shows the difference in MSE (PC-Lasso - LMM-Lasso) broken down by both subpopulation and \pb{``coefficient structure'' is vague; can we describe this better?} coefficient structure. Results are shown for phenotypes generated with $\eta = 0.5$, $\xi = 0.8$, and Independent Subpopulations.

\pb{What do you think about dropping the exponential/dichotomous thing and just going with dichotomous throughout?  Doesn't really seem super important.  Well, we can just keep it in mind and if we need to cut down on space / figures, we can think about it.}

The right-hand panels of Figure \ref{fig:big_small_gamma} are consistent with Figure \ref{fig:big_vs_small}, showing superior LMM-Lasso performance in the presence of fine subpopulation structure, regardless of coefficient magnitude or concordance. The upper left hand panel of Figure \ref{fig:big_small_gamma} corresponds to coarse subpopulation structure with discordant coefficients and also appears to be a setting where LMM-Lasso outperforms PC-Lasso, regardless of coefficient magnitude, although there is some overlap with PC-Lasso. The lower left hand panel of Figure \ref{fig:big_small_gamma} shows results for the setting of coarse supopulation structure with concordant coefficients. In this setting, PC-Lasso and LMM-Lasso perform comparably in the presence of exponential-concordant environmental effects, while LMM-Lasso outperforms PC-Lasso in the majority of dichotomous-concordant environmental effects. These findings align with our initial hypothesis that PC-Lasso will perform well in the presence of simple environmental effect patterns that can be well-approximated by a linear combination of the observed data. When that is not the case, either due to the presence of fine subpopulation structure, dichotomous effects, or genetic similarity that does not reflect similar environmental effects, LMM-Lasso exhibits estimation accuracy superior to that of PC-Lasso.

@@

\pb{Not sure I love this figure, as it emphasizes exp vs dichotomous, which isn't really very interesting.  Can we re-panel it so that it emphasizes discordant vs concordant?}

@@

LMM-Lasso-Rakitsh and LMM-Lasso-ggmix achieve nearly identical MSE for $\bbeta$ in many scenarios, particularly where $n << p$ \pb{you want $n \ll p$, not $n << p$; this occurs a bunch}, as shown in the top panel of Figure \ref{fig:eta_beta_mse}. However, when $n \ge p$ the estimation performance of LMM-ggmix becomes erratic, and in some cases worse than that of LMM-Lasso-Rakitsch. This is most dramatic in the presence of substantial environmental environmental confounding, $\xi = 0.8$, and $n = \{1200, 1400\}$.

LMM-Lasso-ggmix achieves an improvement in the estimation of $\eta$ compared to LMM-Lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n << p$, as shown in the middle panel of Figure \ref{fig:eta_beta_mse}. These improved $\eta$ estimates do not correspond to smaller $\bbeta$ MSE for LMM-Lasso-ggmix and thus do not seem to improve its estimation of the fixed genetic effects. In addition, when $n \ge 1000$, LMM-Lasso-ggmix's estimation of $\eta$ suffers. When $n \ge 1000$ and substantial environmental confounding is present, $\xi = 0.8$,  LMM-Lasso-ggmix's performance is hindered to the point that LMM-Lasso-Rakitsch estimates $\eta$ more accurately.

Overall these results show that LMM-Lasso-ggmix performs as well as LMM-Lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-Lasso-Rakitsch in estimating $\eta$ when $n << p$. However, these are also the settings in which LMM-Lasso-ggmix is least computationally efficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n << p$, LMM-Lasso-ggmix is up to 60 times slower than LMM-Lasso-Rakitsch. Conversely, in settings where $n \ge p$, LMM-Lasso-ggmix is up to 10 times faster than LMM-Lasso-Rakitsch. However, these are also settings in which LMM-Lasso Rakitsch substantially outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ accurately and consistently.

@@

We considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus more broadly applicable.

>>>>

\subsubsection{Subpopulation structure}
\label{sec:subpopulation_structure}
For the fully simulated data, two subpopulation structures were considered, which we refer to as ``fine'' and ``coarse''.  We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, representing a continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations (also known as founding populations) are of equal size. We note that the distinction between fine and coarse structure is an entirely different concept than the level of admixture in the simulated data. It is the size of the subpopulations, not the level of admixture, that determines what we refer to as fine or coarse structure. For example, both the Independent Subpopulation and 1D Linear Admixture simulated data may be comprised of fine or coarse structure. However, the subpopulations will be more distinct in the unadmixed Independent Subpopulation data than in the 1D Linear Admixed data. The Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \ref{fig:admixed} - \ref{fig:empirical} illustrate the simulated kinship structures, as well as that for the Empirical SNP data.

@@

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. This is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by nonoverlapping genetic subpopulations, meaning that a greater proportion of the environmental effect is in the column space of $\boldsymbol{P}_{\bX}$. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:sim_env_conf}. Here, we have CP $= 0.47, 0.99$ for the 1D Linear Admixture and Independent Subpopulations, respectively. The large CP of the Independent Subpopulations model reflects the extreme confounding and estimation bias for this data, for which the PC-Lasso and LMM-Lasso methods are able to correct. Meanwhile, the less dramatic improvement in MSE for the 1D Linear Admixture data by the PC-Lasso and LMM-Lasso methods confirms what we know from this data's substantially smaller CP value; that there is less confounding and bias which can be corrected for.

Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical CP $= 0.61$ with those from the simulated data sets is not straightforward. We note that the CP decreases in the presence of subpopulations of unequal sizes and with more outbred subpopulations as in the Empirical data.

@@

Given the superior performance of PC-Lasso and LMM-Lasso compared to the ordinary Lasso in the presence of environmental confounding, we next compare these methods in the presence of fine and coarse subpopulation structures as defined in section \ref{sec:subpopulation_structure}. Synthetic phenotypes were simulated using a 1:1 SNR, $\eta = 0.5$, and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/mse_diff_subpops}
     \includegraphics[width=0.8\textwidth]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. }
    \label{fig:big_vs_small}
\end{figure}

@@

In addition to evaluating PC-Lasso and LMM-Lasso in the presence of coarse and fine subpopulation structure, we also consider various patterns of environmental effects. Figure \ref{fig:big_small_gamma} shows the difference in MSE (PC-Lasso - LMM-Lasso) broken down by both subpopulation and \anna{environmental effect pattern}. Results are shown for phenotypes generated with $\eta = 0.5$, $\xi = 0.8$, and Independent Subpopulations.

\pb{What do you think about dropping the exponential/dichotomous thing and just going with dichotomous throughout?  Doesn't really seem super important.  Well, we can just keep it in mind and if we need to cut down on space / figures, we can think about it.} \anna{so leave for now?}

\anna{RE-WRITE THIS TO MAKE SURE IT MAKES SENSE WITH NEW FIGURES} The right-hand panels of Figure \ref{fig:big_small_gamma} are consistent with Figure \ref{fig:big_vs_small}, showing superior LMM-Lasso performance in the presence of fine subpopulation structure, regardless of coefficient magnitude or concordance. The upper left hand panel of Figure \ref{fig:big_small_gamma} corresponds to coarse subpopulation structure with discordant coefficients and also appears to be a setting where LMM-Lasso outperforms PC-Lasso, regardless of coefficient magnitude, although there is some overlap with PC-Lasso. The lower left hand panel of Figure \ref{fig:big_small_gamma} shows results for the setting of coarse supopulation structure with concordant coefficients. In this setting, PC-Lasso and LMM-Lasso perform comparably in the presence of exponential-concordant environmental effects, while LMM-Lasso outperforms PC-Lasso in the majority of dichotomous-concordant environmental effects. These findings align with our initial hypothesis that PC-Lasso will perform well in the presence of simple environmental effect patterns that can be well-approximated by a linear combination of the observed data. When that is not the case, either due to the presence of fine subpopulation structure, dichotomous effects, or genetic similarity that does not reflect similar environmental effects, LMM-Lasso exhibits estimation accuracy superior to that of PC-Lasso.

@@

@@

LMM-Lasso-Rakitsh and LMM-Lasso-ggmix achieve nearly identical MSE for $\bbeta$ in many scenarios, particularly where $n \ll p$, as shown in the top panel of Figure \ref{fig:eta_beta_mse}. However, when $n \ge p$ the estimation performance of LMM-ggmix becomes erratic, and in some cases worse than that of LMM-Lasso-Rakitsch. This is most dramatic in the presence of substantial environmental environmental confounding, $\xi = 0.8$, and $n = \{1200, 1400\}$.

LMM-Lasso-ggmix achieves an improvement in the estimation of $\eta$ compared to LMM-Lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n \ll p$, as shown in the middle panel of Figure \ref{fig:eta_beta_mse}. These improved $\eta$ estimates do not correspond to smaller $\bbeta$ MSE for LMM-Lasso-ggmix and thus do not seem to improve its estimation of the fixed genetic effects. In addition, when $n \ge 1000$, LMM-Lasso-ggmix's estimation of $\eta$ suffers. When $n \ge 1000$ and substantial environmental confounding is present, $\xi = 0.8$,  LMM-Lasso-ggmix's performance is hindered to the point that LMM-Lasso-Rakitsch estimates $\eta$ more accurately.

Overall these results show that LMM-Lasso-ggmix performs as well as LMM-Lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-Lasso-Rakitsch in estimating $\eta$ when $n \ll p$. However, these are also the settings in which LMM-Lasso-ggmix is least computationally efficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n \ll p$, LMM-Lasso-ggmix is up to 60 times slower than LMM-Lasso-Rakitsch. Conversely, in settings where $n \ge p$, LMM-Lasso-ggmix is up to 10 times faster than LMM-Lasso-Rakitsch. However, these are also settings in which LMM-Lasso Rakitsch substantially outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ accurately and consistently.

@@

We considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n \ll p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n \ll p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus more broadly applicable.
